Predicting tree failure likelihood for utility risk mitigation via artificial intelligence

### 2021-03-04 docs/manuscript/manuscript.tex
penultimate manuscript revisions

%% Fonts
% \usepackage{fourier}
% \usepackage[T1]{pbsi}
%\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{minted}
\usepackage[figurename=Fig.,labelfont=bf,labelsep=period]{caption}
\usepackage{subcaption}
\usepackage{amsmath}
%\usepackage{rotating}
\usepackage{newtxtext,newtxmath}
\usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
\usepackage[table]{xcolor}

%% Cite Title
%\usepackage[style=numeric,backend=biber,natbib,sorting=none,maxcitenames=2,maxbibnames=99,doi=false,isbn=false,url=false,eprint=false]{biblatex}
%\addbibresource{../ai-trees-references.bib}

%%% Counters
% \usepackage{chngcntr,mathtools}
% \counterwithout{figure}{section}
% \counterwithout{table}{section}

% \numberwithin{equation}{section}

@@

%AMS-TeX packages
%\usepackage{amssymb,amsmath,amsthm}
%\usepackage{bm}
%\usepackage[mathscr]{eucal}
%\usepackage{colortbl}
%\usepackage{color}

%\DeclareRobustCommand{\hlcyan}[1]{{\sethlcolor{cyan}\hl{#1}}}

@@

\usepackage{multirow,array,multicol}

%
%\definecolor{slblue}{rgb}{0,.3,.62}
% \hypersetup{
%     colorlinks,%
%     citecolor=blue,%
%     filecolor=blue,%
%     linkcolor=blue,
%     urlcolor=slblue
% }

%%%TIKZ
% \usepackage{tikz}
% \usepackage{pgfplots}
% \usepackage{pgfplotstable}
% \usepackage{pgfgantt}
% \pgfplotsset{compat=newest}

% \usetikzlibrary{arrows,shapes,positioning}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{shadows,automata}
% \usetikzlibrary{patterns,fit}
% %\usetikzlibrary{circuits.ee.IEC}
% \usetikzlibrary{decorations.text}
% % For Sagnac Picture
% \usetikzlibrary{%
%     decorations.pathreplacing,%
%     decorations.pathmorphing%
% }

% \newcommand{\osn}{\oldstylenums}
% \newcommand{\dg}{^{\circ}}
% \newcommand{\lt}{\left}
% \newcommand{\rt}{\right}
% \newcommand{\pt}{\phantom}
% \newcommand{\tf}{\therefore}
% \newcommand{\?}{\stackrel{?}{=}}
% \newcommand{\fr}{\frac}
% \newcommand{\dfr}{\dfrac}
% %\newcommand{\ul}{\underline}
% \newcommand{\tn}{\tabularnewline}
% \newcommand{\nl}{\newline}
% \newcommand\relph[1]{\mathrel{\phantom{#1}}}
% \newcommand{\cm}{\checkmark}
% \newcommand{\ol}{\overline}
% \newcommand{\rd}{\color{red}}
% \newcommand{\bl}{\color{blue}}
% \newcommand{\pl}{\color{purple}}
% \newcommand{\og}{\color{orange!90!black}}
% \newcommand{\gr}{\color{green!40!black}}
% \newcommand{\nin}{\noindent}
% \newcommand{\la}{\lambda}
% \renewcommand{\th}{\theta}
% \newcommand{\al}{\alpha}
% \newcommand{\G}{\Gamma}
% \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
%             \node[shape=circle,draw,thick,inner sep=1pt] (char) {\small #1};}}

% \newcommand{\p}{\partial}
% \newcommand{\pd}[2]{\frac{\partial{#1}}{\partial{#2}}}
% \newcommand{\dpd}[2]{\dfrac{\partial{#1}}{\partial{#2}}}
% \newcommand{\pdd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\NameTag{Oke, \today}

@@

\begin{abstract}
  Critical to the resilience of utility power lines, tree failure assessments have historically been performed via
  manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
  likelihood categories under four classification scenarios.  Starting with an original set of 505 expert-labeled images, we perform
  preprocessing and augmentation tasks to increase the number of samples. We optimize several hyperparameters in our CNN
  and test its performance for three different image resolutions. The trained classifier has an average validation
  accuracy of at least 0.94 across all scenario-resolution combinations.  Thus, via this novel framework, we
  demonstrate the potential of artificial intelligence to automate and consequently reduce the costs of tree failure
  likelihood assessments, thereby promoting sustainable infrastructure.
\end{abstract}

@@

    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{improbable-example-2}
    \caption{Improbable}
  \end{subfigure}
  \caption{Examples of training images in each of the three tree risk categories considered in this study.
  Trees in the right-hand column were categorized as improbable due to their lack of structural defects as well as good physiological health. Trees in the center column were categorized as possible due to weak branch unions and crown dieback. Trees in the left-hand column were categorized as probable because they were dead. Leaves in the bottom image in the left-hand column are from vines attached to the dead tree. Images are antialiased in this figure for greater clarity}
  \label{fig:raw_images}
\end{figure}

\begin{table}[h!]\small
    \centering
    \begin{tabular}{l r}\toprule
    \bf Category     & \bf Number of images  \\ \midrule
    \textit{Improbable} & 322\\
    \textit{Possible} & 80 \\
    \textit{Probable} & 56 \\\midrule
    Total & 505 \\\bottomrule
    \end{tabular}
    \caption{Category distribution  in the set of raw input images}
    \label{tab:classdist}
\end{table}

@@

  \caption{Selected processed training images under each classification scenario.
    The images shown are processed versions of those shown in \autoref{fig:raw_images}.
    All were randomly cropped along the vertical axis and some horizontally flipped as well.
  }
  \label{fig:processed_images}
\end{figure}

@@

In this paper, we employ a relatively simple CNN architecture, historically inspired by AlexNet \cite{krizhevsky2012imageneta}.
The structure of the CNN is shown in \autoref{fig:cnn} (generated using an automated  framework \cite{bauerle2021net2vis}).
After the input layer (a matrix of pixels from the input image), we use a 64-filter convolutional layer.
The output is downsampled using a pooling layer that returns the maximum ouput from a $2\times 2$ subsample from the previous layer.
Next, we stack two successive convolutional layers each with a depth of 128 filters.
We follow these with a maximum-pooling layer and then two further 256-filter convolutional layers.
A final maximum-pooling layer is used before we ``flatten'' all outputs into a one-dimensional (fully-connected) array of neurons. 
After flattening the outputs, we use a dense layer to reduce the number of outputs a specified number of units.
Batch normalization \cite{ioffe2015batch} is employed after the first dense layer to improve training efficiency.
A second dense layer is used prior to the output layer, with the number of outputs corresponding to the number of classes in the dataset.
A regularization technique known as ''dropout'' is used after each dense layer.
In each dropout layer, a proportion of the neurons are randomly zeroed during training in order to improve the robustness of the model.
\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{net2vis/graph.pdf}

    \medskip

    \includegraphics[width=.7\textwidth]{figures/net2vis-name-labels/legend}
    \caption{Diagram of convolutional neural network structure. Hyperparameters that are optimized include the number of units in the penultimate dense layers.}
    \label{fig:cnn}
\end{figure}

@@

\subsection{Analysis of classification strategies}
Given that there was no significant difference in performance among the three resolutions tested, we focus our attention on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all the four scenarios and then evaluated performance on the following validation metrics: average precision, average recall and accuracy.
%We investigated in detail the performance of the four classification scenarios outlined earlier, using the 128-pixel resolution images.
The goal of this analysis was to explore the efficiencies of various classification strategies for tree failure risk.
Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios. The metrics are shown in \autoref{fig:cnn-perf}.
In each scenario, we trained the model over 20 epochs, but we employed early stopping using validation loss as the criterion with a patience of 3. That is, the training routine would set the weights of the final model at the epoch from which there was no further improvement in validation loss after three subsequent iterations. From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three. This indicates the uncertainty surrounding the subjective risk classification of the trees to begin with. Given the results from the sensitivity tests, however, it is possible that a different training instance may have provided better results, or perhaps, further iterations were required to achieve convergence.

@@

We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix}.
Each row of the confusion matrix indicates the proportion of observations in a given class that are predicted to be in the classes across the columns.
Thus, the diagonal entry in each matrix represents the class-specific recall score, $Re_{c}$.
The matrices show that generally, the ``Possible'' category is difficult to predict accurately as an individual class.
The classifier performed best when it only had to distinguish between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}). Performance only suffered slightly when \textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}).

@@

\eject
\section{Conclusion}
We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood (\textit{Probable}, \textit{Possible} and \textit{Improbable}) with respect to utility infrastructure. Specifically, we developed a convolutional neural network with state-of-the-art configurations.
We applied data augmentation and pre-processing strategies to increase the size of our dataset by a factor of 5, thus generating 2525 images.
From an initial resolution of $4032 \times 3024$ pixels, we created three sets of image resolutions: $64 \times 64$, $128 \times 128$ and $224 \times 224$.
We then optimized eight of our CNN hyperparameters for 12  classification scenario and image resolution combinations. 
We conducted ten model training trials in each of the 12 cases, using a 20\% validation set to measure model performance.
In these trials, the average accuracy of classification was $\geq 0.94$, with a maximum standard deviation of 0.7.
There was no statistically significant difference between the performance across image resolutions.

We further conducted more detailed classification performance analyses over single training instances for the 128-pixel case.
Our results indicate that the CNN performed best at recalling \textit{Probable} vs. \textit{Improbable} cases.
The \textit{Possible} category appeared to be a confounding case.
This was not unexpected, given the uncertainty and subjectivity in assessing these trees to begin with.
The CNN, however, performed better at distinguishing between \textit{Possible}/\textit{Improbable} and \textit{Probable}, than between \textit{Probable}/\textit{Possible} and \textit{Improbable}.
This might be an indicator that trees in the \textit{Possible} category are more likely to be identified as \textit{Improbable}.
When all three failure-likelihood categories were predicted individually, the \textit{Possible} category had the lowest recall score.

Nevertheless, given the relatively small input dataset of original images, these preliminary results are extremely promising for future improvements.
First, we can train better models with more data.
We also plan to rigorously quantify the uncertainty in ground truth category assignments by incorporating predictions from multiple experts on the same images.
In order to better understand how the CNN is classifying each image, we will conduct extensive visual inference in further work, for instance, using the gradient-weighted class activation mapping approach \cite{zeiler2014visualizing}.

@@

\section{Acknowledgments}
The authors acknowledge the support of Eversource.

>>>>

%% Fonts
% \usepackage{fourier}
% \usepackage[T1]{pbsi}
%\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{minted}
\usepackage[figurename=Fig.,labelfont=bf,labelsep=period]{caption}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{newtxtext,newtxmath}
\usepackage[colorlinks=true,citecolor=red,linkcolor=black]{hyperref}
\usepackage[table]{xcolor}

@@

%AMS-TeX packages
%\usepackage{amssymb,amsmath,amsthm}
%\usepackage{bm}
%\usepackage[mathscr]{eucal}
%\usepackage{colortbl}

@@

\usepackage{multirow,array,multicol}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\NameTag{Oke, \today}

@@

\begin{abstract}
  Critical to the resilience of utility power lines, tree failure assessments have historically been performed via
  manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
  likelihood categories (\textit{Probable}, \textit{Possible}, \textit{Improbable}) under four classification scenarios.
  Starting with an original set of 505 expert-labeled images, we perform preprocessing and augmentation tasks to
  increase the number of samples. We optimize several hyperparameters in our CNN and then train and test its performance
  for three different image resolutions. The trained CNN produced a validation accuracy of at least 0.94
  ($\hat\sigma = 0.1)$ in the best-performing, yet hypothetical scenario as it excludes one of the categories.  The
  second-best performing scenario, which includes all categories and therefore more practical, resulted in a validation
  accuracy of 0.92 ($\hat\sigma = 0.1$). Thus, via this novel framework, we demonstrate the potential of artificial
  intelligence to automate and consequently reduce the costs of tree failure likelihood assessments, thereby promoting
  sustainable infrastructure.
\end{abstract}

@@

    \includegraphics[width=\textwidth,keepaspectratio=true,angle=-90]{improbable-example-2}
    \caption{Improbable}
  \end{subfigure}
  \caption{Examples of training images in each of the three tree risk categories considered in this study.  Trees in  
     column (a) were categorized as \textit{Improbable} due to their lack of structural defects as well as good
    physiological health. Trees in the center column were categorized as \textit{Possible} due to weak branch unions and crown
    dieback. Trees in column (c)  were categorized as \textit{Probable} because they were dead. Leaves in the bottom
    image in the left-hand column are from vines attached to the dead tree.}
  \label{fig:raw_images}
\end{figure}

\begin{table}[h!]\small
    \centering
    \begin{tabular}{l r}\toprule
    \bf Category     & \bf Number of images  \\ \midrule
    \textit{Probable} & 56 \\
      \textit{Possible} & 80 \\
      \textit{Improbable} & 322\\\midrule
    Total & 505 \\\bottomrule
    \end{tabular}
    \caption{Category distribution  in the set of raw input images}
    \label{tab:classdist}
\end{table}

@@

  \caption{Selected processed training images under each classification scenario.
    The images shown are processed versions of those shown in \autoref{fig:raw_images}.
    All were randomly cropped along the vertical axis and 50\% were horizontally flipped (including some in this figure).
  }
  \label{fig:processed_images}
\end{figure}

@@

In this paper, we employ a relatively simple CNN architecture, historically inspired by AlexNet \cite{krizhevsky2012imageneta}.
The structure of the CNN is shown in \autoref{fig:cnn} (generated using an automated  framework \cite{bauerle2021net2vis}).
After the input layer (a matrix of pixels from the input image), we use a 64-filter convolutional layer.
The output is downsampled using a pooling layer that returns the maximum output from a $2\times 2$ subsample from the previous layer.
Next, we stack two successive convolutional layers each with a depth of 128 filters.
We follow these with a maximum-pooling layer and then two further 256-filter convolutional layers.
A final maximum-pooling layer is used before we ``flatten'' all outputs into a one-dimensional (fully-connected) array of neurons. 
After flattening the outputs, we use a dense layer to reduce the number of outputs a specified number of units.
Batch normalization \cite{ioffe2015batch} is employed after the first dense layer to improve training efficiency.
A second dense layer is used prior to the output layer, with the number of outputs corresponding to the number of classes in the dataset.
A regularization technique known as ''dropout'' is used after each dense layer.
In each dropout layer, a proportion of the neurons are randomly zeroed during training in order to improve the robustness of the model.
\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{net2vis/graph.pdf}

    \bigskip

    \includegraphics[width=.5\textwidth, trim={4cm 0 0 0}, clip]{figures/net2vis-name-labels/legend}
    \caption{Diagram of convolutional neural network structure (excluding the input layer). Hyperparameters that are
      optimized include the number of units in the penultimate dense layers. Here, 480 and 448 units are used, respectively.}
    \label{fig:cnn}
\end{figure}

@@

Scenario Pr\_Im unsurprisingly has the best performance (greatest accuracy and $Pr^{m}$ of 0.95), since the
classifier only has to predict two extreme categories of \textit{Probable} and \textit{Improbable}.  However, the
\textit{Possible} category cannot be evaded in reality.  Thus, from the perspective of practical application, scenario
PrPo\_Im, which is the next best performing scenario (best accuracy of 0.92, $\hat\sigma=0.01$), demonstrates that it is
most viable to group \textit{Probable} with \textit{Possible} for the best CNN performance.

\subsection{Analysis of classification strategies}
Given that there was no significant difference in performance among the three resolutions tested, we focused our
attention on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all the four
scenarios using this resolution and then evaluated performance on the afore-mentioned validation metrics: accuracy and the macro-averages of
class-specific precision, recall and $F_{1}$ score.
The goal of this analysis was to explore the efficiencies of various classification strategies for tree failure risk.

Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the
respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios. The
metrics are shown in \autoref{fig:cnn-perf}.  In each scenario, we trained the model over 13 epochs. % but we employed
% early stopping using validation loss as the criterion with a patience of 3.
% That is, the training routine would set the
% weights of the final model at the epoch from which there was no further improvement in validation loss after three
% subsequent iterations.
From this figure, we see that Pr\_Po\_Im performs significantly worse than the other three. This indicates the
uncertainty surrounding the expert, yet subjective, failure likelihood assessment of the trees to begin
with, particularly when the category is deemed to be \textit{Possible}. % Given the results
% from the sensitivity tests, however, it is possible that a different training instance may have provided better
% results, or perhaps, further iterations were required to achieve convergence.

@@

We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix} to further investigate the
performance of the classifier.  Each row of the confusion matrix indicates the proportion of observations in a given
class that are predicted to be in the classes across the columns.  Thus, the diagonal entry in each matrix represents
the class-specific recall score, $Re_{c}$.  The matrices show that generally, the \textit{Possible} category is
difficult to predict accurately as an individual class.  The classifier performs best when it only has to distinguish
between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}). Performance only suffers slightly when
\textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}), which we would
select as the best strategy from a practical standpoint.  These two scenarios (Pr\_Im and PrPo\_Im) have the highest
class-specific recall scores $Re_{c}$ ($\ge 0.90$) compared to the other two scenarios Pr\_PoIm ($Re_{c} \ge 0.58$) and
Pr\_Po\_Im ($Re_{c} \ge 0.54)$. In scenario Pr\_PoIm, we see that even grouping \textit{Possible} with
\textit{Improbable} worsens the ability of the classifier to distinguish the \textit{Probable} category.

@@

\eject
\section{Conclusion}
We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood
(\textit{Probable}, \textit{Possible} and \textit{Improbable}) with respect to utility infrastructure. Specifically, we
developed a convolutional neural network with state-of-the-art configurations.  We applied data augmentation and
pre-processing strategies to increase the size of our dataset by a factor of 5, thus generating 2525 images.  From an
initial resolution of $4032 \times 3024$ pixels, we created three sets of image resolutions: $64 \times 64$,
$128 \times 128$ and $224 \times 224$. We then defined four classification scenarios to investigate the performance of
various groupings of the three categories: Pr\_Im (\textit{Probable} vs.\ \textit{Improbable}); PrPo\_Im
(\textit{Probable} and \textit{Possible} vs.\ \textit{Improbable}); Pr\_PoIm (\textit{Probable} vs.\ \textit{Possible}
and \textit{Improbable}); and Pr\_Po\_Im (\textit{Probable} vs.\ \textit{Possible} vs. \textit{Improbable}).
We then optimized eight of our CNN hyperparameters for 12 classification
scenario and image resolution combinations.  We conducted five-fold cross-validation for each of the 12 cases, and
assessed model performance based on accuracy and the macro-averages of precision, recall and $F_{1}$ score.

Our results indicated that there generally was no significant difference in performance between the resolutions.  Among
the scenarios, however, Pr\_Im performed the best with a top accuracy and $Pr^{m}$ of 0.94 ($\hat\sigma = 0.01)$.
The next best-performing scenario was PrPo\_Im with a top accuracy and $Pr^{m}$ of 0.92 ($\hat\sigma = 0.01$).  For
practical applications, scenario PrPo\_Im is more realistic, as it includes all three categories.  Thus, we deemed this
as the most viable.  Scenarios Pr\_PoIm and Pr\_Po\_Im had best accuracy scores of 0.91 ($\hat\sigma = 0.01$) and 0.84
($\hat\sigma = 0.02$), respectively.  But their performance across the other metrics was considerably worse.

We further conducted more detailed classification performance analyses over single training instances for the 128-pixel
case.  Our results indicate that the CNN performed best at recalling \textit{Probable} vs. \textit{Improbable} or
\textit{Probable}/\textit{Possible} vs.\ \textit{Improbable}.  The \textit{Possible} category appeared to be a
confounding for the classifier.  This was not unexpected, given the uncertainty and subjectivity in assessing these trees
to begin with.  The CNN, however, performed better at distinguishing between \textit{Possible}/\textit{Improbable} and
\textit{Probable}, than between \textit{Probable}/\textit{Possible} and \textit{Improbable}.  This might be an indicator
that trees in the \textit{Possible} category are more likely to be identified as \textit{Improbable}.  When all three
failure-likelihood categories were predicted individually in scenario Pr\_Po\_Im, the \textit{Possible} category had the
lowest class-specific recall score.

Nevertheless, given the relatively small input dataset of original images, these preliminary results are extremely
promising for future improvements.  First, we can train better models with more data.  We also plan to rigorously
quantify the uncertainty in ground truth category assignments by incorporating predictions from multiple experts on the
same images.  In order to better understand how the CNN is classifying each image, we will conduct extensive visual
inference in further work, for instance, using the gradient-weighted class activation mapping approach
\cite{zeiler2014visualizing}.  There is also a potential for mapping the visual cues learned by the CNN to physical
relationships governing tree structure, in order to gain greater insights into tree failure processes.  The automation
for tree failure likelihood assessments can potentially supplement human decision-making for increased resilience and,
in the future, reduce costs and improve reliability in tree assessments, thus leading to more sustainable communities.

@@

\section{Acknowledgments}
The authors acknowledge the partial support of Eversource in funding this work.
